\subsection{Merging multiple alignments together}

The \prog{cmalign} program of \textsc{infernal} is capable of merging
two alignments into one. The two alignments must have both been created by
\prog{cmalign} (and the same version of \prog{cmalign}, 1.0 or later),
and must have been created using the same exact CM. This ability is
potentially useful for saving time when aligning a large number of
sequences if you have access to a compute cluster, as described in the
previous section (``Splitting up large alignment jobs''), or if you
want to merge an existing reference alignment with a newly created
one. Combining a new alignment with a reference may be useful in
downstream phylogenetic analysis if you know the classification of the
sequences in the reference alignment.

Imagine we wanted to merge the archaeal sequences from the
\prog{rocks.fa} sequence dataset from the basic tutorial with 
an alignment of the archaeal seed sequences. 

Merging alignments only makes sense and saves time if you've already
computed the two alignments you want to merge. For this example I've
provided the two alignments in the \prog{/tutorial} subdirectory:

\begin{description}
\item[\emprog{rocks.archaea.stk}]
  An alignment of the archaeal sequences from the \prog{rocks.fa}
  dataset. The basic tutorial steps through how to create this file.

\item[\emprog{seed.archaea.stk}]
  An alignment of the archaeal seed sequences, created with the
  command \prog{ssu-align ../seeds/archaea-0p1.cm ../seeds/archaea-0p1.fa seed ../ssualign.0p1.params}
  (Note: this file is NOT identical to the seed alignment
  \prog{archaea-0p1.stk}. I've realigned those sequences to the
  archaeal model to obtain posterior probability confidence estimates
  in the alignment. 
\end{description}

If \prog{cmalign} version 1.0 is in your path, you 
can merge these two alignments into a single alignment called
\prog{seed-rocks.archaea.stk} with

\user{cmalign --merge -o seed-rocks.archaea.stk ../seeds/archaea-0p1.cm seed.archaea.stk rocks.archaea.stk}

The resulting alignment will be 100\% identical to the alignment
\prog{cmalign} would have created if it were used to align a single
sequence file with the seed sequences and rocks sequences together.
